On Optimality of Finite Element Integration

### 2015-09-25 TOMS_Paper/assembly.tex
Revisiting the cost function section

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item the cost function $\theta$ is optimal; that is, it predicts correctly the profitability of pre-evaluation
\item pre-evaluating different monomials does not result in identical tables
\item there is no sharing of terms among monomials in $P$
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}. 2) This is inherently a pathological case, from which we abstract. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We also abstract from this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.

The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2}) = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively. $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.

@@

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true. If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds. Because of assumption 2), the new pre-evaluations cannot expose further sharing. Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.

B) In absence of sharing, the statement is trivially true since we would have $\Lambda''_{ops_2} > \Lambda'_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8). The rationale is that since we would have to pay anyway the presence of $z$ in the innermost loop, adding $p$ to $Z$ would not augment the operation count in our model, so we can safely avoid pre-evaluation.

% assunzione 3 è quella che previene la somma di due mass matrix moltiplicate da F0 una e da F1 l'altra...ma non solo ! chesso magari la trial function è derivata in un monomial e nell'altra no, ma la test function è sempre uguale...
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie one loose hand: the construction of the pre-evaluation cost function $\theta$. We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the sharing-free operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned. Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$. Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Simplified loop nest for the local assembly of a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}

One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$. In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities. By expanding the product we observe an increase in the number of $[L_j, L_k]$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient). Intuitively, $\iota$ captures this growth in non-hoistable operation.

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients expressed over the same function space. The expansion would now lead to identical sub-expressions that, once pre-evaluated, would be mapped to the same temporary. The resulting loop nest would be characterized by sharing, as a result. Therefore, the actual operation count (i.e., once sharing is removed) would be smaller than that one could infer from analysing the expansion ``in isolation''. For a precise estimate of $\iota$, we then need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).

>>>>

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item the cost function $\theta$ is optimal. We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, it returns two values $\theta_{w}$ and $\theta_{wo}$, representing the operation counts with and without pre-evaluation;
\item pre-evaluating different monomials does not result in identical tables;
\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}. 2) This is inherently a pathological case, from which we abstract. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We also abstract from this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.

The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2})$ where $\Lambda'_{ops_1}$ and $\Lambda'_{ops_2}$ represent the operation counts for $P$ and $Z$, respectively. $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.

@@

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if all monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is trivially true. If, on the other hand, there is no (or only partial) sharing, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by pre-evaluating more monomials, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds. The intuition is that this should naturally be true since $\theta$ is optimal (assumption 1) and it predicted pre-evaluation to be a losing strategy for $Z_P$. However, to guarantee this we need to rely on assumption 2): if the new pre-evaluated tables were identical to those resulting from $P$, then we would have more sharing opportunities, which would generally break the optimality claim. Therefore, assumptions 1) and 2) are necessary for A) to be true.

B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$. We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$ (both obvious due to assumption 1), we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. An extreme example is the summation of mass matrices, in which each monomial is characterized by same test and trial functions, but also by a different pre-multiplying coefficient (i.e., known) function. Assumption 3) enables us abstracting from such a scenario. 
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie one loose hand: the definition of the cost function $\theta = [\theta_{w}, \theta_{wo}]$. Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple expression tree traversal suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$. Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}

One can think of this as the (simplified) loop nest originating from the local assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities. By expanding the products, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$). Intuitively, $\iota$ captures the increase in non-hoistable operations due to the transformations that expose pre-evaluation opportunities.

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients over the same function space. Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated. The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that expected by a naive analysis. For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
